---
title: 'Harmonization of Proteomics Data'
author: "Jan-Ole Schulze"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    number_sections: false
    toc: true
    toc_float: true
    toc_depth: 5
---

```{r setup, include=FALSE}
install.packages("remotes")

remotes::install_github("symbod/proharmed-R")
library("proharmed")
```

# What this script does
It adds columns to the protein dataset:
Original Data:              Data after Harmonization:
Protein.Ids        →        Protein.Ids
Genes (messy)      →        Genes (original)
[intensity cols]   →        FilteredProteinIDs        ← NEW
                            RemappedGeneNames         ← NEW  
                            ReducedGeneNames          ← NEW
                            Orthologs                 ← NEW
                            [intensity cols]

Removes ROWS Where:
- Contamination proteins (REV__, CON__) or non-rat proteins in the FilteredProteinIDs step
- Proteins without canonical Ensembl gene names (ReducedGeneName does not exist)
                    


## Run Proharmed Harmonization

In this guide, we'll navigate through the essential steps of harmonizing proteomics data. Our focus is on refining proteomic datasets to enhance their clarity and utility. We begin with filtering protein IDs to ensure data accuracy, followed by gene name remapping and reduction for streamlined dataset representation. We also explore the mapping of orthologs, crucial for cross-species analysis. Utilizing the `proharmed` package, we'll demonstrate these processes and complement them with insightful visualizations, providing a clear view of each step in the data processing journey. This guide is a concise resource for effective proteomics data harmonization.

### Config

```{r config}
prot_file <- "input/report.pg_matrix.tsv"
meta_file <- "input/SampleDescription.csv"

save_plots <- TRUE
# Define the final output directory or set NULL if results should not be saved
final_out_dir <- "harmonization_results"
# Creates the folder if it doesn't exist
if(!is.null(final_out_dir)){dir.create(final_out_dir, recursive = TRUE, showWarnings = FALSE)}
```

### Load Data & Function

```{r load_data}
# Load data
prot_data <- data.table::fread(prot_file)
meta_data <- data.table::fread(meta_file)

head(prot_data)
head(meta_data)
```

## Helper Functions

```{r helper_function, include=FALSE} 
# Function to save each data.frame in the list
save_data_frames <- function(in_list, subdir) {
  for (name in names(in_list)) {
    filepath <- file.path(subdir, paste0(name, ".csv"))
    # Create a copy of the data.frame for modification
    df <- in_list[[name]]
    # Check for columns that are lists and concatenate their elements
    list_cols <- sapply(df, is.list)
    if (any(list_cols)) {
      df[list_cols] <- lapply(df[list_cols], function(col) {
        sapply(col, function(cell) {
          if (is.null(cell)) return(NA)
          paste(cell, collapse = ";")
        })
      })
    }
    write.csv(df, file = filepath, row.names = FALSE)
  }
}
```


### Filtering Protein IDs

The `filter_protein_ids` function is designed to process a DataFrame containing protein IDs. It allows for the removal of decoy IDs (such as those starting with "REV__" or "CON__"), contamination IDs, and the option to filter IDs based on a specified organism. Additionally, the function can restrict protein IDs to reviewed ones and manage empty cells in the dataset.

#### Parameters

- `data`: A DataFrame that includes a column with protein IDs.
- `protein_column`: The name of the column in `data` that contains the protein IDs.
- `organism`: (Optional) Specify the organism to filter the protein IDs. [human, rat, mouse, rabbit]
- `rev_con`: (Optional) A boolean value indicating if decoy protein IDs should be kept.
- `keep_empty`: A boolean value indicating if rows with empty protein ID cells (after filtering!) should be retained.
- `res_column`: The name of the column where the filtered protein IDs will be stored. If `NULL`, the original `protein_column` will be overwritten.
- `reviewed`: (Optional) A boolean value indicating if the protein IDs should be reduced to reviewed ones only.

#### Run Method

```{r filter_proteins}
# Define the parameters
protein_column <- "Protein.Ids"
organism <- "rat"  
rev_con <- FALSE 
keep_empty <- FALSE
res_column <- "FilteredProteinIDs"
reviewed <- FALSE

# Call the function
filtered_prot_results <- proharmed::filter_protein_ids(data = prot_data, protein_column = protein_column, 
                                                       organism = organism, rev_con = rev_con, 
                                                       keep_empty = keep_empty, res_column = res_column,
                                                       reviewed = reviewed)

filtered_prot_data <- filtered_prot_results$Modified_Data # The original dataframe post filtering
```

#### Save Dataframes

```{r filter_proteins_save, include=FALSE}
# Create a subdirectory for filtered_protein_ids
if(!is.null(final_out_dir)){
  subdir <- file.path(final_out_dir, "filtered_protein_ids", "")
  dir.create(subdir, recursive = TRUE, showWarnings = FALSE)
  # Save the data.frames in the list to the specified subdirectory
  save_data_frames(filtered_prot_results, subdir)
}
```


#### Visualizing results

##### Overview Plot

`create_overview_plot` generates an overview plot of the logging data. This plot provides a high-level summary of the data processing steps.

Parameters:

- `logging`: A DataFrame containing the overview logging data returned by one of the proharmed methods.
- `out_dir`: (Optional) The directory where the plot will be saved.
- `file_type`: (Optional) The file format for the saved plot (e.g., png, pdf, jpg).

Returns: A list of two `ggplot` objects.

##### Detailed Plot for Filtered Protein IDs

`create_filter_detailed_plot` creates a detailed plot specifically for the logging data of the `filter_protein_ids` method. This plot provides insights into how the protein IDs were filtered based on various criteria.

Parameters:

- `logging`: A DataFrame containing detailed logging data returned by the `filter_protein_ids` method.
- `organism`: The organism specified in the `filter_protein_ids` method.
- `reviewed`: The reviewed parameter value used in the `filter_protein_ids` method.
- `decoy`: The rev_con parameter value used in the `filter_protein_ids` method.
- `out_dir`: (Optional) The directory where the plot will be saved.
- `file_type`: (Optional) The file format for the saved plot.

Returns: A `ggplot` object.

```{r filter_proteins_plots, message=FALSE, warning=FALSE}
if(save_plots && !is.null(final_out_dir))
  { out_dir=subdir } else { out_dir=NULL }

file_type="png"

library(reshape2)
# Generate and visualize the overview plot
overview_plots <- proharmed::create_overview_plot(logging = filtered_prot_results$Overview_Log, 
                                                  out_dir = out_dir, file_type = file_type)
print(overview_plots[[1]]) # Distribution of number of entries per row
print(overview_plots[[2]]) # Number of entries added and removed

# Generate and visualize the detailed filter plot
detailed_plot <- proharmed::create_filter_detailed_plot(logging = filtered_prot_results$Detailed_Log, 
                                                        organism = organism, reviewed = reviewed, 
                                                        decoy = rev_con, out_dir = out_dir, file_type = file_type)
print(detailed_plot) # Number of entries removed by cause
```

### Remap Gene Names

The `remap_genenames` function in the proharmed package is designed to remap protein IDs to their associated gene names, offering various modes for filling empty entries or optionally replacing existing gene names. It processes the data row by row and performs the remapping of the genes, but the details depend on the configurations.

Modes:

- `all`: Uses primarily information from the FASTA headers, supplemented with UniProt information.
- `fasta`: Utilizes information extracted from FASTA headers.
- `uniprot`: Employs mapping information from UniProt to use all gene names.
- `uniprot_primary`: Uses UniProt mapping information but limits to primary gene names.
- `uniprot_one`: Uses UniProt mapping but only employs the most frequent single gene name.

#### Parameters

- `data`: A DataFrame containing a column with protein IDs.
- `mode`: Mode of refilling (options are all, fasta, uniprot, uniprot_primary, uniprot_one).
- `protein_column`: Name of the column with protein IDs.
- `gene_column`: (Optional) Name of the column with gene names. Defaults to "Gene Names".
- `res_column`: Name of the column for results. If NULL, the `gene_column` will be overwritten.
- `skip_filled`: (Optional) A boolean value indicating if already filled gene name cells should be skipped.
- `organism`: (Optional) Specify the organism for matching IDs. [human, rat, mouse, rabbit]
- `fasta`: (Optional) FASTA file to use when mode is all or fasta.
- `keep_empty`: A boolean value indicating if rows with empty remapped gene names (after remapping!) should be retained.

#### Run Method

```{r remap_gene_names}
# Define the parameters
mode <- "uniprot_one"  # Choose from 'all', 'fasta', 'uniprot', 'uniprot_primary', 'uniprot_one'
protein_column <- "FilteredProteinIDs"
gene_column <- "Genes"
res_column <- "RemappedGeneNames" 
skip_filled <- TRUE # Skip every row where a gene name is provided
organism <- "rat"  
fasta <- NULL  # Specify FASTA file path, if required
keep_empty <- TRUE

remapped_prot_results <- proharmed::remap_genenames(data =  filtered_prot_data, mode = mode, 
                                                    protein_column = protein_column, 
                                                    gene_column = gene_column, 
                                                    res_column = res_column, 
                                                    skip_filled = skip_filled, 
                                                    organism = organism, 
                                                    fasta = fasta, 
                                                    keep_empty = keep_empty)

remapped_prot_data <- remapped_prot_results$Modified_Data
```

#### Save Dataframes

```{r remap_gene_names_save, include=FALSE}
# Create a subdirectory for remap_gene_names
if(!is.null(final_out_dir)){
  subdir <- file.path(final_out_dir, "remap_gene_names", "")
  dir.create(subdir, recursive = TRUE, showWarnings = FALSE)
  # Save the data.frames in the list to the specified subdirectory
  save_data_frames(remapped_prot_results, subdir)
}
```


#### Visualizing results

##### Overview Plot

`create_overview_plot` generates an overview plot of the logging data. This plot provides a high-level summary of the data processing steps.

Parameters:

- `logging`: A DataFrame containing the overview logging data returned by one of the proharmed methods.
- `out_dir`: (Optional) The directory where the plot will be saved.
- `file_type`: (Optional) The file format for the saved plot (e.g., png, pdf, jpg).

Returns: A list of two `ggplot` objects.

```{r remap_gene_names_plots, message=FALSE, warning=FALSE}
if(save_plots && !is.null(final_out_dir))
{ out_dir=subdir } else { out_dir=NULL }

file_type="png"

# Generate and visualize the overview plot
overview_plots <- proharmed::create_overview_plot(logging = remapped_prot_results$Overview_Log, 
                                                  out_dir = out_dir, file_type = file_type)
print(overview_plots[[1]]) # Distribution of number of entries per line
print(overview_plots[[2]]) # Number of entries added and removed
```

### Reduce Gene Names

The `reduce_genenames` function in the `proharmed` package is designed to streamline gene names in a dataset by reducing redundant names to a single representative name. It supports several modes of operation for this reduction process:

- `ensembl`: Reduces gene names to those with an Ensembl ID, using gProfiler.
- `HGNC`: Utilizes the HGNC database for reduction (specific to human genes).
- `mygeneinfo`: Employs the mygene.info database for gene name reduction.
- `enrichment`: Reduces gene names based on functional annotation, using gProfiler.

#### Parameters

- `data`: A DataFrame containing a column with gene names.
- `mode`: The mode of reduction (ensembl, HGNC, mygeneinfo, enrichment).
- `gene_column`: Name of the column with gene names.
- `organism`: The organism for matching IDs.
- `res_column`: Name of the column for results. If NULL, the `gene_column` will be overwritten.
- `keep_empty`: Boolean indicating whether to retain rows with empty reduced gene names.
- `HGNC_mode`: Setting for selecting gene names in HGNC (options: mostfrequent, all). Relevant only for the HGNC mode.

#### Run Method

```{r reduce_gene_names}
# Define the parameters
mode <- "ensembl"  # Choose from 'ensembl', 'HGNC', 'mygeneinfo', 'enrichment'
gene_column <- "RemappedGeneNames"
organism <- "rat" 
res_column <- "ReducedGeneNames"
keep_empty <- FALSE
HGNC_mode <- "mostfrequent"  # 'mostfrequent' or 'all' for HGNC mode

# Reducing gene names
reduced_prot_results <- proharmed::reduce_genenames(data = remapped_prot_data, 
                                                    mode = mode, 
                                                    gene_column = gene_column, 
                                                    organism = organism,
                                                    res_column = res_column, 
                                                    keep_empty = keep_empty, 
                                                    HGNC_mode = HGNC_mode)

reduced_prot_data <- reduced_prot_results$Modified_Data
```


#### Save Dataframes

```{r reduce_gene_names_save, include=FALSE}
# Create a subdirectory for reduce_gene_names
if(!is.null(final_out_dir)){
  subdir <- file.path(final_out_dir, "reduce_gene_names", "")
  dir.create(subdir, recursive = TRUE, showWarnings = FALSE)
  # Save the data.frames in the list to the specified subdirectory
  save_data_frames(reduced_prot_results, subdir)
}
```

#### Visualizing results

##### Overview Plot

`create_overview_plot` generates an overview plot of the logging data. This plot provides a high-level summary of the data processing steps.

Parameters:

- `logging`: A DataFrame containing the overview logging data returned by one of the proharmed methods.
- `out_dir`: (Optional) The directory where the plot will be saved.
- `file_type`: (Optional) The file format for the saved plot (e.g., png, pdf, jpg).

Returns: A list of two `ggplot` objects.

##### Detailed Plot for Reduced Gene Names

`create_reduced_detailed_plot` is a specialized function for generating a detailed visualization of the logging data associated with the `reduce_genenames` method from the proharmed package. This plot offers a granular view of the gene name reduction process, highlighting key steps and outcomes.

Parameters

- `logging`: A DataFrame containing detailed logging data returned by the reduce_genenames method of proharmed.
- `out_dir`: (Optional) The directory where the plot will be saved. If not specified, the plot is not saved but returned for display.
- `file_type`: (Optional) The file format for the saved plot (options include 'png', 'pdf', 'jpg', etc.).

Returns: A `ggplot` object.

```{r reduce_gene_names_plots, message=FALSE, warning=FALSE}
if(save_plots && !is.null(final_out_dir))
{ out_dir=subdir } else { out_dir=NULL }

file_type="png"

# Generate and visualize the overview plot
overview_plots <- proharmed::create_overview_plot(logging = reduced_prot_results$Overview_Log, 
                                                  out_dir = out_dir, file_type = file_type)
print(overview_plots[[1]]) # Distribution of number of entries per row
print(overview_plots[[2]]) # Number of entries added and removed

# Generate and visualize the detailed reduced gene names plot
detailed_plot <- proharmed::create_reduced_detailed_plot(logging = reduced_prot_results$Detailed_Log, 
                                                         out_dir = out_dir, file_type = file_type)
print(detailed_plot) # Number of entries removed by cause
```

### Map Orthologs

The `map_orthologs` function from the `proharmed` package is designed to map gene names from one organism to their orthologs in another organism. This function is particularly useful for cross-species analyses and comparative genomics studies.

#### Parameters

- `data`: A DataFrame containing a column with gene names.
- `gene_column`: The name of the column in data that contains the gene names.
- `organism`: The organism corresponding to the current gene names in the data.
- `tar_organism`: The target organism to which you want to map the orthologs.
- `res_column`: The name of the column where the orthologs will be stored. If NULL, the gene_column will be overwritten.
- `keep_empty`: A boolean value indicating whether rows with empty cells (where orthologs could not be found) should be kept or deleted.

#### Run Method

```{r map_orthologs}
# Define the parameters
gene_column <- "ReducedGeneNames"
organism <- "rat" 
tar_organism <- "human"
res_column <- "Orthologs"  # Column name for the results
keep_empty <- TRUE 

# Mapping orthologs
orthologs_prot_results <- proharmed::map_orthologs(data = reduced_prot_data, gene_column = gene_column, 
                                                   organism = organism, tar_organism = tar_organism, 
                                                   res_column = res_column, keep_empty = keep_empty)

orthologs_prot_data <- orthologs_prot_results$Modified_Data
```

#### Save Dataframes

```{r map_orthologs_save, include=FALSE}
# Create a subdirectory for map_orthologs
if(!is.null(final_out_dir)){
  subdir <- file.path(final_out_dir, "map_orthologs", "")
  dir.create(subdir, recursive = TRUE, showWarnings = FALSE)
  # Save the data.frames in the list to the specified subdirectory
  save_data_frames(orthologs_prot_results, subdir)
}
```

#### Visualizing results

##### Overview Plot

`create_overview_plot` generates an overview plot of the logging data. This plot provides a high-level summary of the data processing steps.

Parameters:

- `logging`: A DataFrame containing the overview logging data returned by one of the proharmed methods.
- `out_dir`: (Optional) The directory where the plot will be saved.
- `file_type`: (Optional) The file format for the saved plot (e.g., png, pdf, jpg).

Returns: A list of two `ggplot` objects.

##### Detailed Plot for Gene Mapping

`create_ortholog_detailed_plot` is a tailored function for creating a detailed visualization of the logging data generated by the `map_orthologs` method in the proharmed package. This function produces a comprehensive plot that illustrates the intricate process of mapping orthologs from one organism to another, showcasing the critical steps and results achieved during this procedure.

Parameters

- `logging`: A DataFrame containing detailed logging data returned by the `map_orthologs` method of proharmed.
- `organism`: The source organism used in the `map_orthologs` method.
- `out_dir`: (Optional) The directory where the plot will be saved. If not specified, the plot is not saved but returned for display.
- `file_type`: (Optional) The file format for the saved plot (options include 'png', 'pdf', 'jpg', etc.).

Returns: A `ggplot` object.

```{r map_orthologs_plots, message=FALSE, warning=FALSE}
if(save_plots && !is.null(final_out_dir))
{ out_dir=subdir } else { out_dir=NULL }

file_type="png"

# Generate and visualize the overview plot
overview_plots <- proharmed::create_overview_plot(logging = orthologs_prot_results$Overview_Log, 
                                                  out_dir = out_dir, file_type = file_type)
print(overview_plots[[1]]) # Distribution of number of entries per line
print(overview_plots[[2]]) # Number of entries added and removed - removed genes could not be mapped to orthologs, they were not actually removed

# Generate and visualize the detailed ortholog mapping plot
detailed_plot <- proharmed::create_ortholog_detailed_plot(logging = orthologs_prot_results$Detailed_Log, 
                                                          organism = organism, 
                                                          out_dir = out_dir, file_type = file_type)
print(detailed_plot) # Number of entries removed by cause
```
```{r}
# Final summary at the end of the script
cat("\n=== HARMONIZATION SUMMARY ===\n")
cat("Initial dataset:", nrow(prot_data), "proteins\n")
cat("After filtering (removed non-rat and contaminants):", nrow(filtered_prot_data), "proteins\n")
cat("After gene name remapping:", nrow(remapped_prot_data), "proteins\n")
cat("After Ensembl reduction (removed proteins without canonical IDs):", nrow(reduced_prot_data), "proteins\n")
cat("After ortholog mapping:", nrow(orthologs_prot_data), "proteins\n")
cat("\nTotal proteins removed:", nrow(prot_data) - nrow(orthologs_prot_data), "\n")

cat("=== COLUMN NAME MATCHING CHECK ===\n")
for (frac in unique(meta_data$fraction)) {
  frac_samples <- meta_data$sample_name[meta_data$fraction == frac]
  found_in_data <- intersect(frac_samples, names(orthologs_prot_data))
  not_found <- setdiff(frac_samples, names(orthologs_prot_data))
  
  cat(sprintf("\n%s fraction:\n", frac))
  cat(sprintf("  Samples in metadata: %d\n", length(frac_samples)))
  cat(sprintf("  Samples found in data: %d\n", length(found_in_data)))
  cat(sprintf("  Samples NOT found in data: %d\n", length(not_found)))
  if(length(not_found) > 0) print(not_found)
}
```

```{r}
cat("=== PROTEIN LOSS AT EACH HARMONIZATION STEP ===\n")

steps <- list(
  "Initial"            = prot_data,
  "Filter protein IDs" = filtered_prot_data,
  "Remap gene names"   = remapped_prot_data,
  "Reduce gene names"  = reduced_prot_data,
  "Map orthologs"      = orthologs_prot_data
)

count_per_fraction <- function(data, meta) {
  counts <- c()
  for (frac in c("AI", "AS")) {
    frac_samples <- intersect(meta$sample_name[meta$fraction == frac], names(data))
    if (is.data.table(data)) {
      subset <- data[, frac_samples, with = FALSE]
    } else {
      subset <- data[, frac_samples, drop = FALSE]
    }
    has_value <- rowSums(!is.na(subset)) > 0
    counts[frac] <- sum(has_value)
  }
  return(counts)
}

# Print header
cat(sprintf("%-25s %8s %8s\n", "Step", "AI", "AS"))
cat(strrep("-", 45), "\n")

prev_counts <- NULL
for (name in names(steps)) {
  counts <- count_per_fraction(steps[[name]], meta_data)
  
  if (is.null(prev_counts)) {
    cat(sprintf("%-25s %8d %8d\n", name, counts["AI"], counts["AS"]))
  } else {
    cat(sprintf("%-25s %8d %8d    (AI: -%d, AS: -%d)\n",
                name,
                counts["AI"], counts["AS"],
                prev_counts["AI"] - counts["AI"],
                prev_counts["AS"] - counts["AS"]))
  }
  prev_counts <- counts
}

cat(strrep("-", 45), "\n")
initial_counts <- count_per_fraction(prot_data, meta_data)
final_counts <- count_per_fraction(orthologs_prot_data, meta_data)
cat(sprintf("%-25s %8d %8d\n", "Total removed",
            initial_counts["AI"] - final_counts["AI"],
            initial_counts["AS"] - final_counts["AS"]))

cat("=== ZERO CHECK IN SAMPLE COLUMNS ===\n")
sample_cols <- intersect(meta_data$sample_name, names(prot_data))

total_zeros <- sum(prot_data[, ..sample_cols] == 0, na.rm = TRUE)
cat(sprintf("Total zeros in sample columns: %d\n", total_zeros))
```
