---
title: 'DE Analysis of Proteomics Data'
author: "Jan-Ole Schulze"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    number_sections: false
    toc: true
    toc_float: true
    toc_depth: 5
---

```{r setup, include=FALSE}
library_names <- c("PRONE", "dplyr", "data.table", "SummarizedExperiment", "tidyr", "ComplexUpset", "ggplot2")

for (lib in library_names) {
  library(lib, character.only = TRUE)
}
```

# Config
```{r config}
fraction_to_analyze <- "AS"  # Set to "AI" or "AS"

primary_norm_method <- "RobNorm_scaled"

# Additional normalization, run alongside the primary one purely as a diagnostic
diagnostic_norm_methods <- c("LoessF_scaled")

norm_dir  <- file.path(paste0("normalization_results_", fraction_to_analyze), "scaled")
prot_file <- file.path(norm_dir, paste0(primary_norm_method, "_normalized_data.csv"))
meta_file <- file.path(norm_dir, "meta_data.csv")

save_plots <- TRUE

final_out_dir <- paste0("de_analysis_results_", fraction_to_analyze, "_", primary_norm_method)
# Creates the folder if it doesn't exist
if(!is.null(final_out_dir)){dir.create(final_out_dir, recursive = TRUE, showWarnings = FALSE)}
```

## Prepare Data
```{r load_data}
prot_data <- data.table::fread(prot_file) 
meta_data <- data.table::fread(meta_file)
```

## Load into Summarized Experiment
In this section, we utilize the `load_data` function to import and structure proteomics data for further analysis. This function is specifically designed to handle real-world proteomics datasets, transforming them into a `SummarizedExperiment` object, which is a standard format for handling complex experimental data in bioinformatics.

**The load_data Function**
The `load_data` function efficiently loads and organizes proteomics data, taking into account various aspects of the experimental design:
- `data`: This is the main dataset with proteins as rows and samples as columns (e.g., proteinGroups.txt from MaxQuant).
- `md`: The metadata or experimental design table, which should include a "Column" that corresponds to sample intensity column names in the data.
- `protein_column`: The column name in data containing protein IDs.
- `gene_column`: The column name in data containing gene names.
- `ref_samples`: Reference samples, particularly useful in TMT experiments.
- `batch_column`: The column in md that defines experimental batches.
- `condition_column`: The column in md that defines experimental conditions.
- `label_column`: A column in md containing simple sample names for visualization purposes.

**Preparing and Loading Data**
In the following R code chunk, we first subset our metadata (`md`) to the samples that actually have a column in the protein data. We then call the load_data function, passing our protein data (`prot_data`), the subset metadata, and specifying relevant columns for protein IDs, gene names, reference samples, batches, conditions, and labels.

```{r prepare_data}
md <- meta_data[meta_data$Column %in% names(prot_data)]

se <- load_data(data = prot_data, md = md, 
                protein_column = "Protein.IDs", gene_column = "Gene.Names", 
                ref_samples = NULL, 
                batch_column = NULL, condition_column = NULL, label_column = "Label")

# Remove the "log2" assay that is automatically created, because the data is already log-transformed
assays(se)[["log2"]] <- NULL
# Rename the "raw" assay to "normalized"
names(assays(se))[names(assays(se)) == "raw"] <- primary_norm_method

# Attach the diagnostic assays, matched on Protein.IDs rather than assumed to be
# row-aligned.
for (m in diagnostic_norm_methods) {
  d <- data.table::fread(file.path(norm_dir, paste0(m, "_normalized_data.csv")))
  assay(se, m, FALSE) <- as.matrix(d[match(rowData(se)$Protein.IDs, d$Protein.IDs),
                                     colnames(se), with = FALSE])
}

cat("Available assays in the SummarizedExperiment object:\n")
print(names(assays(se)))
```

# Differential Expression Analysis

## Prepare Data
In this section, we prepare our dataset for analysis, crucial for comparing different experimental conditions and timepoints in biological research.
- We extract and modify column data from our `SummarizedExperiment` object (`se`), creating a new column `Cond_Scaffold_Time`. This column uniquely identifies each condition, scaffold and timepoint by combining them with an underscore.
- The updated column data is then reassigned back to `se`, ensuring our dataset includes these detailed identifiers.
- The comparisons themselves are then listed explicitly in the next chunk, as pairs of `Cond_Scaffold_Time` values. This list is vital for subsequent comparative analyses, allowing for systematic examination across various experimental conditions.

```{r}
# Inspect your metadata to find all available columns
cat("=== METADATA COLUMN NAMES ===\n")
print(names(meta_data))

cat("\n=== COLDATA COLUMN NAMES (after SE creation) ===\n")
print(names(colData(se)))

col_data <- colData(se)

col_data$Cond_Scaffold_Time <- paste(col_data$Condition, 
                                     col_data$Scaffold, 
                                     col_data$Timepoint,
                                     sep = "_")

# Print all unique values in the new column
cat("\nAll unique Cond_Scaffold_Time values:\n")
print(unique(col_data$Cond_Scaffold_Time))

colData(se) <- col_data
```

## Define comparisons
```{r}
# Configure comparisons of interest
selected_comparisons <- c(
  "diabetic_empty_42-nondiabetic_empty_42",
  "diabetic_PCL_42-nondiabetic_PCL_42"
)

# Display names for figures
comparison_labels <- c(
  "diabetic_empty_42-nondiabetic_empty_42" = "Empty bone defect",
  "diabetic_PCL_42-nondiabetic_PCL_42"     = "PCL scaffold"
)
```

# Run DE Analysis with Protein Filtering 
The `run_DE` function performs DE analysis on the proteomics data within our `SummarizedExperiment` object (`se`). This function is adeptly suited for proteomics data and includes several important parameters:

**Main Parameters**
- `se`: Our `SummarizedExperiment` object, containing the normalized proteomics dataset and relevant metadata.
- `comparisons`: A vector of specified condition pairs for analysis, aimed at detecting significant expression changes between these pairs.
- `DE_method`: The method for DE analysis.

**Additional Parameters Explained**
- `logFC`: A Boolean parameter indicating whether to apply a log fold change (logFC) threshold. If `TRUE`, only proteins with changes exceeding the specified logFC thresholds are considered significant.
- `logFC_up` and `logFC_down`: These specify the upper and lower thresholds for log2 fold change, defining upregulated and downregulated proteins, respectively.
- `p_adj`: A Boolean parameter indicating whether to use adjusted p-values (`TRUE`) or raw p-values (`FALSE`) for significance thresholding.
- `alpha`: The threshold for significance, applied to adjusted or raw p-values based on p_adj.
- `B` and `K`: Parameters specific to the ROTS method in DE analysis. `B` is the number of bootstraps, enhancing the robustness of the analysis, while `K` represents the number of top-ranked features considered for optimizing reproducibility.

The Limma design matrix will look like this, because R automatically expands the single column:
         diabetic_empty_42  nondiabetic_empty_42  diabetic_PCL_42  ...
Sample1         1                   0                      0            ...
Sample2         1                   0                      0            ...
Sample3         0                   1                      0            ...
Sample4         0                   1                      0            ...
Sample5         0                   0                      1            ...
Sample6         0                   0                      1            ...
=> Limma will fit a linear model with one input feature for every of these groups

```{r}
# Function to filter proteins with sufficient non-NA values per group
filter_proteins_by_group <- function(se, comparison, condition_column, min_samples = 3, assay_name = NULL) {
  # Parse comparison string to get the two groups
  groups <- strsplit(comparison, "-")[[1]]
  group1 <- groups[1]
  group2 <- groups[2]
  
  # Get sample metadata
  col_data <- colData(se)
  
  # Find samples belonging to each group
  group1_samples <- rownames(col_data)[col_data[[condition_column]] == group1]
  group2_samples <- rownames(col_data)[col_data[[condition_column]] == group2]
  
  cat("Comparison:", comparison, "\n")
  cat("Group 1 (", group1, "):", length(group1_samples), "samples\n")
  cat("Group 2 (", group2, "):", length(group2_samples), "samples\n")
  
  # Get the assay data
  assay_data <- assays(se)[[assay_name]]
  
  # Convert to matrix if it's a data.table (common issue)
  if (is(assay_data, "data.table")) {
    assay_data <- as.matrix(assay_data)
  }
  
  # Ensure we have the right column names
  available_cols <- colnames(assay_data)
  group1_samples <- intersect(group1_samples, available_cols)
  group2_samples <- intersect(group2_samples, available_cols)
  
  cat("Available group 1 samples:", length(group1_samples), "\n")
  cat("Available group 2 samples:", length(group2_samples), "\n")
  
  if (length(group1_samples) == 0 || length(group2_samples) == 0) {
    stop("No matching samples found for one or both groups. Check sample names and condition column.")
  }
  
  # Count non-NA values per protein in each group
  group1_counts <- rowSums(!is.na(assay_data[, group1_samples, drop = FALSE]))
  group2_counts <- rowSums(!is.na(assay_data[, group2_samples, drop = FALSE]))
  
  # Filter proteins with sufficient data in both groups - Creates TRUE/FALSE vector for each protein
  valid_proteins <- (group1_counts >= min_samples) & (group2_counts >= min_samples)
  
  cat("Proteins before filtering:", nrow(se), "\n")
  cat("Proteins after filtering:", sum(valid_proteins), "\n")
  cat("Proteins removed:", sum(!valid_proteins), "\n\n")
  
  # Return filtered SummarizedExperiment
  return(se[valid_proteins, ])
}

# Function to run DE analysis with pre-filtering
# NOTE: Protein filtering is performed independently for each comparison
# A protein is included in analysis if it has ≥3 non-NA values in BOTH groups
# This means different comparisons may test different numbers of proteins
# `extra_assays` are additional assays limma is also run on, for the diagnostic only.
run_DE_with_filtering <- function(se, comparisons, condition_column, min_samples = 3,
                                  assay_to_use = NULL, extra_assays = character(0)) {
  all_results <- data.frame()
  filtered_proteins_per_comparison <- list()
  
  for (comparison in comparisons) {
    cat("=== Processing comparison:", comparison, "===\n")
    
    # Filter proteins for this specific comparison
    se_filtered <- filter_proteins_by_group(se, comparison, condition_column, min_samples, assay_to_use)
    
    # Store the actual protein IDs that were retained (not row indices!)
    # Get protein IDs from the rowData of the SummarizedExperiment
    if ("Protein.IDs" %in% names(rowData(se_filtered))) {
      filtered_proteins_per_comparison[[comparison]] <- rowData(se_filtered)$Protein.IDs
    }
    
    # Run DE analysis on filtered data
    de_result <- run_DE(se = se_filtered, 
                       comparisons = comparison,  # Single comparison
                       ain = c(assay_to_use, extra_assays),
                       condition = condition_column,
                       DE_method = "limma",
                       logFC = TRUE, 
                       logFC_up = 1,
                       logFC_down = -1, 
                       p_adj = TRUE, 
                       alpha = 0.05
                       )
    
    # Combine results
    all_results <- rbind(all_results, de_result)
  }
  
  return(list(results = all_results, filtered_proteins = filtered_proteins_per_comparison))
}

de_results_container <- run_DE_with_filtering(
  se = se, 
  comparisons = selected_comparisons,
  condition_column = "Cond_Scaffold_Time",
  min_samples = 3,
  assay_to_use = primary_norm_method,
  extra_assays = diagnostic_norm_methods
)

# Extract results and filtered protein info
de_results_all <- de_results_container$results
filtered_proteins_per_comparison <- de_results_container$filtered_proteins

# Everything downstream sees the primary assay alone
de_results <- de_results_all[de_results_all$Assay == primary_norm_method, ]
```

## Visualize Results
The `plot_volcano_DE` function is designed to create informative volcano plots from the DE results:
- `de_res`: This is the data table obtained from the `run_DE` function, containing detailed DE results for each protein in the study.
- `ain`: A vector specifying the normalization methods to visualize. These should be methods already applied in the DE analysis and present in `de_res`.
- `comparisons`: The specific comparisons for which we want to generate volcano plots. These should align with the comparisons used in the DE analysis.
- `facet_norm`: A Boolean parameter that decides if the plots should be faceted (i.e., split into separate panels) by normalization method. Setting this to `FALSE` leads to a more consolidated view.
- `facet_comparison`: Another Boolean parameter that controls whether to facet by comparison. This is particularly useful when facet_norm is `FALSE`, allowing us to compare different condition pairs side-by-side.

```{r}
# Create the volcano plot
volcano_plot <- plot_volcano_DE(de_results, 
                               ain = c(primary_norm_method), 
                               comparisons = selected_comparisons, 
                               facet_norm = FALSE, 
                               facet_comparison = TRUE)

# Check what was returned
cat("Class of volcano_plot:", class(volcano_plot), "\n")
cat("Length of volcano_plot:", length(volcano_plot), "\n")

# If it's a list, save each plot individually
if(is.list(volcano_plot)) {
  for(i in seq_along(volcano_plot)) {
    plot_name <- names(volcano_plot)[i]
    if(is.null(plot_name)) plot_name <- paste0("plot_", i)
    
    ggsave(filename = file.path(final_out_dir, paste0("volcano_plot_DE_", plot_name, ".png")), 
           plot = volcano_plot[[i]], 
           width = 12, 
           height = 8, 
           dpi = 300)
  }
} else {
  # If it's a single plot, save normally
  ggsave(filename = file.path(final_out_dir, "volcano_plot_DE.png"), 
         plot = volcano_plot, 
         width = 12, 
         height = 8, 
         dpi = 300)
}

# 1. Examine the actual logFC values in your data
print("=== LogFC Summary ===")
summary(de_results$logFC)
print(paste("Range:", round(min(de_results$logFC, na.rm=TRUE), 2), "to", round(max(de_results$logFC, na.rm=TRUE), 2)))
```


### Save Results
We save the raw differential expression results as `de_results_raw.csv` for record-keeping and future reference. The results are then merged with the human ortholog annotations carried by the input data, creating a comprehensive dataset named `de_results_with_orthologs`. This enriched dataset is also preserved in a CSV file, `de_results_raw_with_orthologs.csv`, ensuring all critical data and insights are securely stored and accessible for subsequent analysis or review.

```{r}
# Save de_results to de_results_raw.csv
write.csv(de_results, file.path(final_out_dir, "de_results_raw.csv"), row.names = FALSE)

cat("Column names in de_results:\n")
print(names(de_results))

# Extract orthologs from the original protein data
orthologs_data <- prot_data[, c("Protein.IDs", "Gene.Names", "Orthologs")]

# Merge orthologs information with raw DE results
de_results_with_orthologs <- merge(de_results, 
                                  orthologs_data, 
                                  by = c("Protein.IDs", "Gene.Names"), 
                                  all.x = TRUE)

# Save the enhanced raw results
write.csv(de_results_with_orthologs, file.path(final_out_dir, "de_results_raw_with_orthologs.csv"), row.names = FALSE)
```

## Get DEPs
We next focus on isolating differentially expressed proteins (DEPs) from our analysis results. By filtering the DE results, we identify proteins with significant expression changes, categorizing them as DEPs. These are proteins where the expression change is notable and not classified as "No Change." After isolating these significant results, we save them as `de_results_significant.csv`. This step is crucial for honing in on the most impactful proteins that show significant differential expression, which are potential candidates for further biological investigation or validation studies.

### Get Significant DEPs and save results
```{r}
de_results_sig <- de_results[de_results$Change != "No Change"]

write.csv(de_results_sig, file.path(final_out_dir, "de_results_significant.csv"), row.names = FALSE)
```

```{r}
# Count total significant DEPs
total_sig_deps <- nrow(de_results_sig)
cat("=== SUMMARY OF DIFFERENTIALLY EXPRESSED PROTEINS ===\n\n")
cat("Total significant DEPs across all comparisons:", total_sig_deps, "\n\n")

# Count DEPs per comparison
cat("DEPs per comparison:\n")
deps_per_comparison <- table(de_results_sig$Comparison)
print(deps_per_comparison)
cat("\n")
```

## Normalization diagnostic
Placed last so that a failure here cannot withhold the results saved above.

```{r normalization_diagnostic, warning=FALSE, message=FALSE}
# The protein filter is derived from the primary assay alone, so the assays are
# only comparable if they share its missing-value pattern. Where they do not,
# limma silently returns NA for the proteins it cannot estimate in the other
# assay, PRONE counts those as "No Change", and the overlap below comes out too
# low with nothing to show for it.
for (m in diagnostic_norm_methods) {
  n_mismatch <- sum(is.na(as.matrix(assays(se)[[m]])) !=
                    is.na(as.matrix(assays(se)[[primary_norm_method]])))
  if (n_mismatch > 0)
    stop(sprintf("Assay '%s' differs from '%s' in %d missing values, so their DE ",
                 m, primary_norm_method, n_mismatch),
         "results are not computed on the same proteins and cannot be compared.")
}

p_jaccard <- plot_jaccard_heatmap(de_results_all,
                                  ain = c(primary_norm_method, diagnostic_norm_methods),
                                  comparisons = selected_comparisons, plot_type = "facet_comp") +
  ggplot2::scale_fill_gradient(low = "white", high = "#0072B2", limits = c(0, 1)) +
  ggplot2::facet_wrap(~Comparison, scales = "free",
                      labeller = ggplot2::as_labeller(comparison_labels))
print(p_jaccard)
if (save_plots) {
  ggsave(file.path(final_out_dir, "normalization_diagnostic_jaccard.png"),
         p_jaccard, width = 8, height = 4)
}

print(p_jaccard$data)

# Jaccard measures set overlap only. A protein could be shared yet called in
# opposite directions, which overlap alone would hide.
sig <- de_results_all[de_results_all$Change != "No Change", ]
cat("Proteins called in opposite directions by the two normalizations:",
    sum(tapply(sig$Change, paste(sig$Comparison, sig$Protein.IDs),
               function(x) length(unique(x)) > 1)), "\n")
```