---
title: 'Normalization of Proteomics Data'
author: "Jan-Ole Schulze"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    number_sections: false
    toc: true
    toc_float: true
    toc_depth: 5
---

```{r setup, include=FALSE}
# ── Version check against renv.lock ──────────────────────────────────────────
# Run renv::restore() if any package version has drifted from the locked state.
if (file.exists("renv.lock")) {
  lock <- jsonlite::fromJSON("renv.lock")$Packages
  key_pkgs <- c("PRONE", "POMA", "limma", "SummarizedExperiment")
  for (pkg in key_pkgs) {
    if (!is.null(lock[[pkg]])) {
      installed <- tryCatch(as.character(packageVersion(pkg)), error = function(e) "NOT INSTALLED")
      locked    <- lock[[pkg]]$Version
      if (installed != locked)
        warning(sprintf("Package %s: installed=%s, locked=%s — run renv::restore() to realign.",
                        pkg, installed, locked))
    }
  }
}

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if (!requireNamespace("PRONE",       quietly = TRUE)) BiocManager::install("PRONE")
if (!requireNamespace("POMA",        quietly = TRUE)) BiocManager::install("POMA")
library("PRONE")
library("POMA")

library("data.table")
library("ggplot2")
library("SummarizedExperiment")

# Set random number seed
seed = 27
set.seed(seed)
```

## Config
```{r config}
prot_file <- "harmonization_results/map_orthologs/Modified_Data.csv"
meta_file <- "input/SampleDescription.csv"
fraction_to_keep <- "AS"  # Set to "AI" or "AS"

save_plots <- TRUE
# Define the final output directory
final_out_dir <- paste0("normalization_results_", fraction_to_keep)
# Creates the folder if it doesn't exist
if(!is.null(final_out_dir)){dir.create(final_out_dir, recursive = TRUE, showWarnings = FALSE)}
```

### Helper Functions
```{r methods, include=FALSE}
save_png <- function(p, filename) {
  png(filename = filename)
  print(p)
  dev.off()
}
```

## Prepare Data
```{r load_data}
# Load data
prot_data <- data.table::fread(prot_file)
meta_data <- data.table::fread(meta_file)

cat("First few columns in prot_data:\n")
print(head(names(prot_data), 10))
```

##### Changes in Meta Data
- Creating a Copy of `meta_data`:
We start by creating a copy of the original `meta_data` data table and assign it to md. This approach preserves the original `meta_data` for any future references.
We modify and generate column names.
This results in a refined version of md with a focus on the columns "Column", "Batch", "Animal", "Fraction", "Scaffold", "Intervention", "Timepoint", "Tissue", "Condition", and "Label".

```{r change_meta}
# Make copy of meta_data
md <- meta_data

md[, `:=` (
    Label = paste0(animal_no, "_", fraction, "_", genotype, "_", scaffold),
    Animal = animal_no,
    Condition = genotype,
    Fraction = fraction,
    Timepoint = time,
    Batch = study,
    Column = sample_name,
    Tissue = tissue,
    Scaffold = scaffold,
    Intervention = intervention
)]

md <- md[, .SD, .SDcols = c("Column", "Batch", "Animal", "Fraction", "Scaffold", "Intervention", "Timepoint", "Tissue", "Condition", "Label")]
```

##### Changes in Proteomics Data
Handle Zeros: Convert zeros to `NA` in key columns listed in `md$Column`. This is crucial because zeros in proteomics data can indicate missing information, and treating them as `NA` ensures more accurate statistical interpretations.
```{r change_proteomics_data}
# Replace all 0s with NAs in specified columns of prot_data
prot_data[, (md$Column) := lapply(.SD, function(x) { ifelse(x == 0, NA, x) }), .SDcols = md$Column]
# Check for any remaining zeros in the specified columns
remaining_zeros <- sum(sapply(prot_data[, md$Column, with = FALSE], function(col) sum(col == 0, na.rm = TRUE)))
cat("Remaining zeros in specified columns:", remaining_zeros, "\n")
```

### Check protein occurences for AS vs. AI across every sample
```{r}
cat("Cleaned protein data dimensions:", dim(prot_data), "\n")

# Get column indices in protein data for each sample type using cleaned metadata
ai_sample_names <- md$Column[md$Fraction == "AI"]
as_sample_names <- md$Column[md$Fraction == "AS"]

# Initialize counters
proteins_ai_only <- 0
proteins_as_only <- 0
proteins_both <- 0
proteins_none <- 0

# Collect the AS-exclusive identifiers, which are listed individually below
proteins_as_only_list <- c()

# Only analyze proteins with a valid name
valid_proteins <- !is.na(prot_data$ReducedGeneNames) & prot_data$ReducedGeneNames != ""
analysis_data <- prot_data[valid_proteins, ]

# Count proteins by their occurrence pattern
for (i in 1:nrow(analysis_data)) {
  # Check if protein has any non-NA values in each sample type (since zeros were converted to NA)
  has_ai <- any(!is.na(analysis_data[i, ai_sample_names, with=FALSE]), na.rm=TRUE)
  has_as <- any(!is.na(analysis_data[i, as_sample_names, with=FALSE]), na.rm=TRUE)
  
  # Get protein identifier
  protein_id <- analysis_data$ReducedGeneNames[i]
  
  # Categorize based on occurrence
  if (has_ai && has_as) {
    proteins_both <- proteins_both + 1
  } else if (has_ai) {
    proteins_ai_only <- proteins_ai_only + 1
  } else if (has_as) {
    proteins_as_only <- proteins_as_only + 1
    proteins_as_only_list <- c(proteins_as_only_list, protein_id)
  } else {
    proteins_none <- proteins_none + 1
  }
}

# Print results
cat("Proteins found only in AI samples:", proteins_ai_only, "\n")
cat("Proteins found only in AS samples:", proteins_as_only, "\n")
cat("Proteins found in both sample types:", proteins_both, "\n")
cat("Proteins not found in either sample type:", proteins_none, "\n")
cat("Total analyzed:", nrow(analysis_data), "\n")

# Print the list of proteins ONLY in AS samples
cat("\n=== PROTEINS FOUND ONLY IN AS SAMPLES ===\n")
if(length(proteins_as_only_list) > 0) {
  for(i in 1:length(proteins_as_only_list)) {
    cat(sprintf("%d. %s\n", i, proteins_as_only_list[i]))
  }
} else {
  cat("No proteins found exclusively in AS samples.\n")
}
```

### Remove rows belonging to a specified sample fraction
```{r}
### Remove samples based on fraction_to_keep
fraction_to_remove <- ifelse(fraction_to_keep == "AI", "AS", "AI")

cat(sprintf("=== KEEPING %s SAMPLES, REMOVING %s SAMPLES ===\n", 
            fraction_to_keep, fraction_to_remove))

# Print initial state
cat("=== BEFORE FRACTION SAMPLE REMOVAL ===\n")
cat("--- Meta data ---\n")
md_original_shape <- dim(md)
cat("Original metadata shape: ", md_original_shape[1], "rows ×", md_original_shape[2], "columns\n")
cat("AS samples in metadata:", sum(md$Fraction == "AS"), "\n")
cat("AI samples in metadata:", sum(md$Fraction == "AI"), "\n")

cat("\n--- Protein dataset ---\n")
original_prot_shape <- dim(prot_data)
cat("Original protein dataset shape: ", original_prot_shape[1], "rows ×", original_prot_shape[2], "columns\n")

# Step 1: Identify removable sample column names from metadata BEFORE filtering
removable_samples_in_md <- md$Column[md$Fraction == fraction_to_remove]
cat("\nFound", length(removable_samples_in_md), "fraction samples in metadata to remove\n")

# Step 2: Remove the samples of the discarded fraction from the metadata first
cat("\n--- Removing Fraction samples from metadata ---\n")
md_filtered <- md[md$Fraction != fraction_to_remove, ]
md_new_shape <- dim(md_filtered)
cat("New metadata shape: ", md_new_shape[1], "rows ×", md_new_shape[2], "columns\n")
cat("Metadata rows removed:", md_original_shape[1] - md_new_shape[1], "\n")
# Update metadata reference
md <- md_filtered

# Step 3: Remove corresponding samples from protein dataset
cat("\n--- Removing fraction samples from protein dataset ---\n")
# Find the discarded fraction's sample columns that exist in protein data
removable_samples_in_data <- intersect(removable_samples_in_md, names(prot_data))
cat("Corresponding columns found in protein data:", length(removable_samples_in_data), "\n")

if(length(removable_samples_in_data) > 0) {
  # Print some example column names for verification
  cat("Example fraction column to remove:", head(removable_samples_in_data, 1), "\n")
  
  # Create vector of columns to keep
  columns_to_keep <- setdiff(names(prot_data), removable_samples_in_data)
  
  # Filter protein data
  prot_data <- prot_data[, ..columns_to_keep]
} else {
  cat("No corresponding sample columns found in protein data - no filtering needed\n")
}

# Print final state
cat("\n=== AFTER FRACTION SAMPLE REMOVAL ===\n")
cat("--- Meta data ---\n")
cat("Final metadata shape: ", dim(md)[1], "rows ×", dim(md)[2], "columns\n")
cat("Remaining AI samples in metadata:", sum(md$Fraction == "AI"), "\n")
cat("Remaining AS samples in metadata:", sum(md$Fraction == "AS"), "\n")

cat("\n--- Protein dataset ---\n")
new_prot_shape <- dim(prot_data)
cat("Final protein dataset shape: ", new_prot_shape[1], "rows ×", new_prot_shape[2], "columns\n")
cat("Columns removed from protein data:", original_prot_shape[2] - new_prot_shape[2], "\n")

# Verify no columns of the discarded fraction remain in protein data
remaining_removable_cols_in_data <- intersect(removable_samples_in_md, names(prot_data))
cat("Remaining removable columns in protein data:", length(remaining_removable_cols_in_data), "\n")

# Final verification - check consistency
cat("\n=== FINAL CONSISTENCY CHECK ===\n")
kept_columns_in_md <- md$Column[md$Fraction == fraction_to_keep]
kept_columns_in_data <- intersect(kept_columns_in_md, names(prot_data))
cat("Kept samples in metadata:", length(kept_columns_in_md), "\n")
cat("Kept sample columns in protein data:", length(kept_columns_in_data), "\n")
cat("Metadata and protein data consistency:", 
    ifelse(length(kept_columns_in_md) == length(kept_columns_in_data), "✓ CONSISTENT", "✗ INCONSISTENT"), "\n")
```


### Load into Summarized Experiment
In this section, we utilize the `load_data` function to import and structure proteomics data for further analysis. This function is specifically designed to handle real-world proteomics datasets, transforming them into a `SummarizedExperiment` object, which is a standard format for handling complex experimental data in bioinformatics.

SummarizedExperiment Object
├── assays          - The actual data matrices
│   ├── counts      - e.g., raw protein intensities
│   ├── log2        - e.g., log2-transformed values
│   └── normalized  - e.g., normalized values
│
├── rowData         - Information about features (rows)
│   ├── Protein.Ids
│   ├── Gene.Names
│   └── other protein annotations
│
├── colData         - Information about samples (columns)
│   ├── Animal
│   ├── Condition
│   ├── Batch
│   ├── Timepoint
│   └── other sample metadata
│
└── metadata        - Experiment-level information
    └── any additional info about the experiment

**The load_data Function**
The `load_data` function efficiently loads and organizes proteomics data, taking into account various aspects of the experimental design:
- `data`: This is the main dataset with proteins as rows and samples as columns (e.g., proteinGroups.txt from MaxQuant).
- `md`: The metadata or experimental design table, which should include a "Column" that corresponds to sample intensity column names in the data.
- `protein_column`: The column name in data containing protein IDs.
- `gene_column`: The column name in data containing gene names.
- `ref_samples`: Reference samples, particularly useful in TMT experiments.
- `batch_column`: The column in md that defines experimental batches.
- `condition_column`: The column in md that defines experimental conditions.
- `label_column`: A column in md containing simple sample names for visualization purposes.

```{r prepare_data}
se <- load_data(data = prot_data, md = md, 
                protein_column = "Protein.Ids", gene_column = "ReducedGeneNames", # Use this gene column to have exactly one canonical name for the same gene!
                ref_samples = NULL, 
                batch_column = NULL, condition_column = NULL, label_column = "Label")

print(names(assays(se)))
```

### Filter Proteins
In the subsequent step, we refine the proteomics dataset by removing proteins that have no measurable value at all:
- `filter_out_complete_NA_proteins`: This function targets proteins that exhibit missing values across all samples. Removing these proteins ensures that the dataset focuses only on proteins with measurable values, thereby enhancing the quality and reliability of the data. Reverse hits and potential contaminants have already been removed upstream, in the harmonization step.
```{r filter_proteins}
# remove proteins with NAs in all samples
se <- filter_out_complete_NA_proteins(se)
```

### Overview of the Data
In this section, we focus on obtaining a preliminary overview of the proteomics dataset encapsulated within a SummarizedExperiment object. This overview is crucial for understanding the data quality and distribution across various conditions and batches.
- `get_NA_overview`: This function analyzes the extent of missing values (NAs) in the dataset. It returns a list containing the total number of values, the count of NAs, and the percentage of missing values. The analysis is performed on a specified data type, defaulting to log2-transformed values.
- `plot_condition_overview`: This function generates a bar plot visualizing the number of samples per specified condition or batch. The visual representation helps in understanding the distribution of samples across different experimental conditions or batches.

```{r pre_overview}
get_NA_overview(se, ain = "log2")

p1 <- plot_condition_overview(se, condition = "Condition")
p2 <- plot_heatmap(se = se, ain = "log2", color_by = c("Condition", "Scaffold"), label_by = "Animal")[["log2"]]

file_type <- "png"

if(save_plots && !is.null(final_out_dir)){
  sub_dir <- paste0(file.path(final_out_dir,"initial_data_overview"))
  dir.create(sub_dir, recursive = TRUE, showWarnings = FALSE)
   
  ggsave(filename = file.path(sub_dir, paste0("num_per_condition.", file_type)), plot = p1)
  save_png(p = p2, filename = file.path(sub_dir, paste0("heatmap.", file_type)))
}

for (p in list(p1, p2)){
  print(p)
}
```

### Check NAs
After filtering the proteomics dataset, analyzing the pattern of missing values (NAs) is essential for understanding the data's quality and guiding further processing. We employ three distinct plotting functions, each providing a unique perspective on the NA distribution.
- `plot_NA_density`: Generates a density plot to visualize the intensity distribution of proteins, segregating them based on the presence or absence of NAs. This plot helps in identifying any potential biases in data due to missing values.
- `plot_NA_heatmap`: This function crafts a heatmap that represents the pattern of NAs in the dataset. Key customization options include:
    - `color_by`: Specifies the column to color the samples. Set to NULL by default, which uses the condition column of the SummarizedExperiment object. Setting it to "No" omits the color bar.
    - `label_by`: Determines the column for labeling the samples. Defaults to NULL, using the labels column of the SummarizedExperiment object. "No" indicates no sample labeling.
    - `cluster_samples`: Boolean to decide if samples should be clustered.
    - `cluster_proteins`: Boolean to decide if proteins should be clustered.
    - `show_row_dend`: Boolean to control the display of the row dendrogram.
    - `show_column_dend`: Boolean to control the display of the column dendrogram.

```{r filter_proteins_plot, warning=FALSE}
# check NA pattern
p1 <- plot_NA_density(se)
p2 <- plot_NA_heatmap(se, label_by = "No", color_by = "No", 
                      cluster_samples = TRUE, cluster_proteins = TRUE, 
                      show_row_dend = TRUE, show_column_dend = FALSE)

if(save_plots && !is.null(final_out_dir)){
  sub_dir <- paste0(file.path(final_out_dir,"initial_data_overview"))
  dir.create(sub_dir, recursive = TRUE, showWarnings = FALSE)

  ggsave(filename = file.path(sub_dir, paste0("na_density.", file_type)), plot = p1)
  save_png(filename = file.path(sub_dir, "na_heatmap.png"), p = p2)
}

for (p in list(p1, p2)){
  print(p)
}
```

## Pre-normalization visualization
```{r}
p_pre_box_condition <- plot_boxplots(se, ain = "log2", label_by = "Condition", color_by = "Scaffold")

p_pre_pca_cond <- plot_PCA(se, ain = "log2", color_by = "Scaffold", 
                  label_by = "No", shape_by = "Condition")

# Save plots if saving is enabled
if(save_plots && !is.null(final_out_dir)){
  sub_dir <- paste0(file.path(final_out_dir,"initial_data_overview"))
  dir.create(sub_dir, recursive = TRUE, showWarnings = FALSE)
  
  save_png(filename = file.path(sub_dir, "p_pre_box_condition"), p = p_pre_box_condition) 
  save_png(filename = file.path(sub_dir, "pre_pca_cond"), p = p_pre_pca_cond)
}

# Display the plots
print(p_pre_box_condition)
print(p_pre_pca_cond)
```

# Perform Normalization
```{r normalize}
# Remove raw assay since normalization should work on log2 data
assays(se)[["raw"]] <- NULL
print(names(assays(se)))

# See what normalization methods are actually available
available_methods <- get_normalization_methods()
print(available_methods)

# Define normalization methods
normalization_methods <- c("RobNorm", "LoessF")

# Normalize the SummarizedExperiment object
# Normalizes all assays in the object - which should just be the log2 one at this point
se_normalized <- normalize_se(se, methods = normalization_methods)

if (!is.null(final_out_dir)){
  export_data(se = se_normalized, out_dir = final_out_dir, ain = normalization_methods)
}
```

### Qualitative Evaluation
In this section, we employ various visualization techniques to scrutinize the normalized proteomics data encapsulated within a `SummarizedExperiment` object. These visualizations help us understand data distribution and relationships among various experimental factors.
- `Boxplots`: The `plot_boxplots` function visualizes the distribution of normalized data as boxplots. It provides insights into data variability and outliers across samples.
    - `se`: The normalized `SummarizedExperiment` object.
    - `ain`: A vector of strings indicating which assays to use as input. If `NULL`, all normalized assays are plotted side by side.
    - `color_by`: A string specifying the metadata column to use for coloring the samples. If `NULL`, colors are assigned based on conditions.
    - `label_by`: A string specifying the metadata column for labeling samples. If `NULL`, labels are not displayed.
    - `ncol`: Number of columns for faceting the plot.
    - `facet_norm`: A boolean to indicate whether to facet the plot by normalization methods.
- `Density Plots`: The `plot_densities` function produces density plots, which illustrate the distribution of normalized data.
    - `se`: The `SummarizedExperiment` object containing normalized data.
    - `ain`: A vector of strings indicating which assays to use. If `NULL`, all normalized assays are considered.
    - `color_by`: A string specifying the column to color the samples. If `NULL`, colors are based on conditions.
    - `facet_norm`: A boolean specifying whether to facet by normalization methods.
- `PCA Plots`: The plot_PCA function creates PCA plots, which are useful for visualizing data variation and relationships in a reduced dimensionality space.
    - `se`: The normalized SummarizedExperiment object.
    - `ain`: A vector of strings indicating which assays to use. If NULL, all assays are plotted.
    - `color_by`: A string specifying the column for coloring samples. If NULL, colors are based on timepoints.
    - `label_by`: A string specifying the column for labeling samples. If NULL, labels are not displayed.
    - `shape_by`: A string specifying the column for shaping samples. If NULL, shapes are based on conditions.
    - `facet_norm`: A boolean specifying whether to facet the plot by normalization methods.
    - `facet_by`: A string specifying the column for faceting the samples.
    - `ellipse`: A boolean indicating if ellipses should be drawn around clusters.
    - `ncol`: Number of columns for faceting the plot.

```{r qualitative}
p1 <- plot_boxplots(se_normalized, ain = normalization_methods, 
                    color_by = "Scaffold", label_by = "Condition")

p2 <- plot_PCA(se_normalized, ain = normalization_methods, color_by = "Scaffold", 
               label_by = "No", shape_by = "Condition", facet_norm = TRUE)


if(save_plots && !is.null(final_out_dir)){
  sub_dir <- paste0(file.path(final_out_dir,"post_normalization_pre_scaling"))
  dir.create(sub_dir, recursive = TRUE, showWarnings = FALSE)
  
  ggsave(filename = file.path(sub_dir, paste0("boxplot_condition.", file_type)), plot = p1)
  ggsave(filename = file.path(sub_dir, paste0("pca_condition.", file_type)), plot = p2)
}

for (p in list(p1, p2)){
  print(p)
}
```

### Print heatmap after normalization and before scaling
```{r}
p_heat_post_norm <- plot_heatmap(se = se_normalized, ain = "RobNorm", color_by = c("Condition", "Scaffold"), label_by = "Animal")

if(save_plots && !is.null(final_out_dir)){
  sub_dir <- paste0(file.path(final_out_dir,"post_normalization_pre_scaling"))
  dir.create(sub_dir, recursive = TRUE, showWarnings = FALSE)
  
  save_png(filename = file.path(sub_dir, "post_norm_pre_scaling_heatmap.png"), p = p_heat_post_norm)
}

print(p_heat_post_norm)
```

## Scaling
```{r}
scale_se <- function(se_data, methods){

  # Get initial count
  initial_samples <- ncol(se_data)
  cat(sprintf("Initial number of samples: %d\n", initial_samples))

  # Convert column data from SummarizedExperiment to a data table for manipulation
  coldata <- as.data.table(colData(se_data))

  # Iterate only over the intended normalized assays (e.g. "log2" must not be
  # scaled and dropped here - it is still needed elsewhere as-is), regardless
  # of what other assays happen to be present on se_data
  for(ain in methods){
    
    # Convert the current assay to a data frame for processing
    dt <- as.data.frame(assays(se_data)[[ain]])
    
    # For ZDF_validation batch - calculate condition-specific medians of ctrl_42 samples
    # Identify the sample IDs
    diabetic_ctrl42_cols <- coldata[coldata$Batch == "ZDF_validation" & 
                                   coldata$Tissue == "ctrl_42" & 
                                   coldata$Condition == "diabetic", ]$Column
    
    nondiabetic_ctrl42_cols <- coldata[coldata$Batch == "ZDF_validation" & 
                                      coldata$Tissue == "ctrl_42" & 
                                      coldata$Condition == "nondiabetic", ]$Column
    
    # Calculate condition-specific medians
    diabetic_ctrl42_medians <- NULL
    nondiabetic_ctrl42_medians <- NULL
    
    if (length(diabetic_ctrl42_cols) > 0) {
      diabetic_ctrl42_medians <- apply(dt[, diabetic_ctrl42_cols, drop = FALSE], 1, median, na.rm = TRUE) # Calculate median row-wise
    }
    
    if (length(nondiabetic_ctrl42_cols) > 0) {
      nondiabetic_ctrl42_medians <- apply(dt[, nondiabetic_ctrl42_cols, drop = FALSE], 1, median, na.rm = TRUE)
    }
    
    diabetic_valid <- !is.na(diabetic_ctrl42_medians) & 
                 !is.nan(diabetic_ctrl42_medians) & 
                 !is.infinite(diabetic_ctrl42_medians)
    
    nondiabetic_valid <- !is.na(nondiabetic_ctrl42_medians) & 
                        !is.nan(nondiabetic_ctrl42_medians) & 
                        !is.infinite(nondiabetic_ctrl42_medians)
    
    cat(sprintf("\nScaling statistics for assay '%s':\n", ain))
    cat(sprintf("  Proteins with valid diabetic ctrl_42 median: %d / %d (%.1f%%)\n", 
                sum(diabetic_valid), length(diabetic_valid), 
                100 * sum(diabetic_valid) / length(diabetic_valid)))
    cat(sprintf("  Proteins with valid nondiabetic ctrl_42 median: %d / %d (%.1f%%)\n", 
                sum(nondiabetic_valid), length(nondiabetic_valid), 
                100 * sum(nondiabetic_valid) / length(nondiabetic_valid)))

    # Iterate over each unique animal in the experiment
    for (animal in unique(coldata$Animal)){
      
      # Get the batch for this animal
      animal_batch <- unique(coldata[coldata$Animal == animal, ]$Batch)
      if ("ZDF_validation" %in% animal_batch) {
        # Get the condition for this animal
        animal_condition <- unique(coldata[coldata$Animal == animal, ]$Condition)
        
        # Get all samples for this animal in ZDF_validation batch
        r <- coldata[(coldata$Animal == animal) & (coldata$Batch == "ZDF_validation"),]$Column
        
        # Scale samples by condition-specific median
        if (length(r) > 0) {
          if (animal_condition == "diabetic" && !is.null(diabetic_ctrl42_medians)) {
            for (col in r) {
              dt[, col] <- dt[, col] - diabetic_ctrl42_medians
            }
          } else if (animal_condition == "nondiabetic" && !is.null(nondiabetic_ctrl42_medians)) {
            for (col in r) {
              dt[, col] <- dt[, col] - nondiabetic_ctrl42_medians
            }
          }
        }
      } else {
                stop(sprintf("ERROR: Animal %s is from batch '%s', but scaling is only implemented for 'ZDF_validation' batch. 
                     \nPlease check your data or implement scaling for other batches.", 
                     animal, paste(animal_batch, collapse = ", ")))
      }
    }

    # Add the scaled data as a new assay in the SummarizedExperiment object
    assay(se_data, paste0(ain, "_scaled"), FALSE) <- as.data.table(dt)

    # Remove the original (unscaled) assay data
    assay(se_data, ain, FALSE) <- NULL
  }
  
  # Count control samples
  ctrl42_count <- sum(se_data$Tissue == "ctrl_42")
  cat(sprintf("Control samples to be removed: %d ctrl_42 samples\n", ctrl42_count))

  # Remove the ctrl_42 reference samples from the SummarizedExperiment
  samples_to_keep <- se_data$Tissue != "ctrl_42"
  se_scaled <- se_data[, samples_to_keep]
  
  # Report final counts
  final_samples <- ncol(se_scaled)
  cat(sprintf("Final number of samples after filtering: %d\n", final_samples))
  cat(sprintf("Total samples removed: %d\n", initial_samples - final_samples))

  return(se_scaled)
}
```


```{r perform and visualize scaling}
se_scaled <- scale_se(se_normalized, normalization_methods)

if (!is.null(final_out_dir)){
  # Create the scaled assay names (they now have "_scaled" suffix)
  scaled_methods <- paste0(normalization_methods, "_scaled")
  # CREATE THE SUBDIRECTORIES
  scaled_dir <- file.path(final_out_dir, "scaled")
  dir.create(scaled_dir, recursive = TRUE, showWarnings = FALSE)
  # Export scaled data
  export_data(se = se_scaled, out_dir = scaled_dir, ain = scaled_methods)
}

p1 <- plot_boxplots(se_scaled, ain = paste(normalization_methods, "_scaled", sep=""),
                    color_by = "Scaffold", label_by = "Condition")

p2 <- plot_PCA(se_scaled, ain = paste(normalization_methods, "_scaled", sep=""), color_by = "Scaffold", 
               label_by = "No", shape_by = "Condition", facet_norm = TRUE)

if(save_plots && !is.null(final_out_dir)){
  sub_dir <- paste0(file.path(final_out_dir,"post_scaling"))
  dir.create(sub_dir, recursive = TRUE, showWarnings = FALSE)
  
  ggsave(filename = file.path(sub_dir, paste0("boxplot_condition.", file_type)), plot = p1)
  ggsave(filename = file.path(sub_dir, paste0("pca_condition.", file_type)), plot = p2)
}

for (p in list(p1, p2)){
  print(p)
}
```

## Choosing the best normalization based evaluation above.
```{r selection_of_norm}
desired_norm <- "RobNorm_scaled"
```

## Detect outliers
```{r}
# Create SE object with only RobNorm data for outlier detection
se_robnorm_only <- se_scaled
assays(se_robnorm_only) <- assays(se_robnorm_only)[desired_norm]  # Keep only desired dataset

# Now run outlier detection on normalized data
outliers_result_robnorm <- POMA::PomaOutliers(se_robnorm_only,
                                      method = "euclidean", 
                                      outcome = "Condition",
                                      labels = FALSE)

print(outliers_result_robnorm$distance_boxplot)
outlier_samples_robnorm <- outliers_result_robnorm$outliers
print(outlier_samples_robnorm)

outliers_result_robnorm <- POMA::PomaOutliers(se_robnorm_only,
                                      method = "euclidean", 
                                      outcome = "Scaffold",
                                      labels = FALSE)

print(outliers_result_robnorm$distance_boxplot)
outlier_samples_robnorm <- outliers_result_robnorm$outliers
print(outlier_samples_robnorm)
```

## Print heatmap of final dataset
```{r}
p <- plot_heatmap(se = se_scaled, ain = desired_norm, color_by = c("Condition", "Scaffold"), label_by = "No")

if(save_plots && !is.null(final_out_dir)){
  sub_dir <- paste0(file.path(final_out_dir,"post_scaling"))
  dir.create(sub_dir, recursive = TRUE, showWarnings = FALSE)
  
  save_png(filename = file.path(sub_dir, "post_scaling_heatmap.png"), p = p)
}
print(p)
```